### 2018-03-25 paper/sexSelectionSpacing-ver1.tex
Minor edits

Parents' spacing between births has long served as a measure of son preference 
\citep{Leung1988}.
Before prenatal sex determination became available, the only recourse for 
parents who wanted a son---but did not yet have one---was to have the next 
birth sooner.
Son preference is therefore often associated with shorter spacing after the 
births of girls than boys 
\citep{Das1987,Rahman1993,Pong1994,Haughton1996,Arnold1997,Soest2012,Rossi2015}.
Shorter spacing is, in turn, associated with worse health outcomes for girls 
and mothers 
\citep{arnold98,Conde-Agudelo2000,Whitworth2002,Razzaque2005,Rutstein2005,Conde-Agudelo2006}.%
\footnote{
Parents are also more likely to cease childbearing after the birth of 
a son than after a daughter 
\citep{ben-porath76b,Das1987,Arnold1997,clark00,filmer09,Basu2010,Altindag2016}.
}

@@

% [Why India?]
I apply the method to birth histories for Hindu women, using data from 
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2016. 
India is a particularly compelling case.
On the one hand, India has seen dramatic increases in the males-to-females ratio 
at birth over the last three decades as access to prenatal sex determination 
expanded 
\citep{das_gupta97,Sudha1999,Arnold2002,retherford03b,jha06}.%
\footnote{
India is not alone; both China and South Korea saw
significant changes in the sex ratio at birth over the same period 
\citep{Yi1993,park95}.
}
On the other hand, research suggests that son preference in 
India, when measured as ideally having more boys than girls, is decreasing 
over time and with higher education \citep{bhat03,pande07}.
% %
% \footnote{
% This measure of son preference is commonly used in the literature. 
% See, for example, \citet{clark00}, \citet{Jensen2009}, and \cite{Hu2015}.
% }
There have also been substantial changes in access and legality of
prenatal sex determination in India over the period covered.
Abortion has been legal in India since 1971 and still is, but the 
first reports of prenatal sex determination did
not appear until 1982--83 \citep{Sudha1999,bhat06,Grover2006}.
The number of clinics quickly increased, and knowledge about sex selection 
became widespread after a senior government official's wife aborted a 
fetus that turned out to be male \citep[p.\ 598]{Sudha1999}.
In 1994, the Central Government passed the Prenatal Diagnostic Techniques 
(PNDT) Act, making determining and communicating the sex of a fetus illegal.%
\footnote{
Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
The number of convictions has been low.
It took until January 2008 for the first state, Haryana, to reach five convictions.
Hence, private clinics apparently operate with little risk of legal action 
\citep{Sudha1999}.
Furthermore, there is little evidence that bans like this significantly
affect sex ratios \citep{Das-Gupta2016}.
}
Hence, the data make it possible to show how the spacing between births and
sex ratios have changed with the introduction of prenatal sex determination,
and whether the ban affected the relationship between birth spacing and sex 
ratios.

[Outline of findings here]

% [EXPLAIN RESULTS HERE]
% Main findings!
% There are three main findings.
% 
% There is, however, still no evidence of sex selection being used on the first birth.
% 
% Secondly, sex selection appears to be used for securing one son, rather than a large 
% number of sons.
% There is only limited use of sex-selective abortions for better-educated
% women with one or more sons.
% The exception is rural women with one son and one daughter, presumably to compensate for 
% the higher child mortality in rural areas.
% These results are in line with the differential stopping behavior observed in many studies
% before sex selection became available \citep{repetto72,arnold98,dreze01}.
%

@@

As mentioned above, 10 to 20 percent of women gave birth less than nine months
after they were married, and the first spell, therefore, begins at the 
month of marriage.
The exception is for women married very young, where I use the month they 
turned 12.
The second and subsequent spells begin nine months after the previous birth 
because that is the earliest we should expect to observe a new birth.
A few women report births that occurred less than nine months 
after the previous birth; I drop those women from the sample.

All spells continue until either a child is born or censoring occurs.
Censoring can happen for three reasons:
the survey takes place;
sterilization of the woman or her husband;
or too few births are observed for the method to work.
For spell 1 I set censoring at 72 months after marriage,
while for spells 2 and 3 censoring is at 72 months and for spell 4
57 months after a birth can occur.

I group spells into four time periods based on spell start date:
1972--1984, 1985--1994, 1995--2004, and 2005--2016.
These years follow the main changes in availability and legality of 
prenatal sex determination discussed in the Introduction.
Some spells cover two periods because I classify periods based on the 
spell's beginning year.
A couple may, for example, be married in 1984, but not have their first 
child until 1986.
That couple's first spell will be in the 1972--1984 period, even though 
most of the interval falls in the 1985--1994 period.
Hence, prenatal sex determination was likely available when some
children from spells that began in the 1972--1984 period were conceived,
which could result in evidence of sex-selective abortions even for this
period.
Similarly, a spell that started in the 1985--1994 period may have been partly 
or mostly under the PDNT act.
The effect is a downward bias in the differences between the periods.

@@

The area of residence is a dummy variable for the household living in
an urban area [TK check on this for NFHS-4].%
\footnote{
NFHS uses four categories for the area of residence: Large city, small city, town,
and countryside.
To reach a sufficient sample size urban areas are merged into one group.
}
The cost of children is higher in urban areas than in rural areas, and 
there is easier access to prenatal sex determination in urban areas.
Both factors are likely to lead to higher use of sex-selective abortions
in urban than in rural areas.

@@

The predicted sex ratio is the weighted average over the women in the sample's 
individual predicted sex ratios, using the individual probabilities of a birth by 
the end of the spell as weights.
Each woman's predicted sex ratio is the weighted average of the predicted
percentage boys over the months in the spell calculated using equation 
(\ref{eq:probability_son}) and the probability of giving birth in each month 
as weights.%
\footnote{
Imagine a spell has two periods and that the estimated percentage boys for a
woman are 54\% and 66\% and that the likelihood of exiting the spell is 20\% and 40\%
(in other words, there is 40\% chance that she does not have a birth by the end of 
the spell).
This woman' percentage boys is then $\frac{54*0.2+66*0.4}{0.2+0.4} = 62$.
}
The predicted sex ratio captures the percent boys that will be born to women 
in the sample once child bearing for that spell is over.

@@

The results so far show that as sex ratios became more
male-biased the spacing when a woman has only girls became 
longer relative to other sex compositions, presumably because
of increased use of sex selection rather than a reduction in son preference.
Neither the median spacing or the conditional survival curves
can, however, provide us with information on how the use of
sex selection vary \emph{within} a spell.
Furthermore, using conditional survival curves forces everybody 
to finish the spell in the same month and does not provide
any information about how many births there are.
We can, however, learn something about both changes in 
sex selection within a spell and the likelihood of exiting
with a birth using standard  survival curves.
The predicted sex ratio may change within a spell for, at
least, two reasons.

@@

First, parents may stop using prenatal sex determination 
once the duration from the previous birth becomes long enough
because they prefer not to have their children too far apart.
The desire for short spacing could, for example, arise from 
economies of scale in childbearing and rearing that would be lost 
if children were born too far apart \citep[p 374]{schultz97}. 
In this case we should observe the predicted sex ratio decline
with spell length.

Second, if the degree of son preference across women within the sample
but behavior is not affex2cted by duration from the previous birth,
we should observe an increasing share of births come from 
parents with stronger son preference as duration from the
previous birth increases.
Whereas the early births are a combination of births to
people with lower son preference and those with stronger son
preference who conceived a son in the first or second pregnancy,
later births are then predominately from those with stronger
son preference who have not yet conceived a son.
Hence, in this case we should see an increase in the sex ratio 
as duration increases.

@@

At the other extreme, women with eight or more years of education have
experienced an almost complete reversal of their spacing patterns.
Rather than having the next birth sooner as they mostly did before 
sex selection became available, women with either no or one son now 
have substantially longer spacing than if they have two or more sons.
That these changes are likely due to increased use of sex selection 
rather than changes in son preference as shown by the increasingly
significant deviations in the sex ratio from the natural sex ratio.

@@

This points to two important, interconnected, questions for 
future research:
What is the connection between falling fertility and the 
use of sex selection and is the use of sex selection increasing 
or decreasing over time?
It is clear that fertility has been declining over time in India,
although it is important to note that the decline might be
overstated exactly because of an increased use of sex selection and
the resulting longer spacing between births makes it less likely 
that we will observe births to women, even though they eventually
will happen.
The facts that biased sex ratio occur at earlier spells over
time and starting to show up for women with less education 
all indicate that lower fertility and use of sex selection are
related, but a more detailed analysis is needed.
Furthermore, although examining spells separately, as done here, is a useful
first step in understanding how sex selection changes over time,
it does not lend itself easily to examine the degree to which
sex selection changes over time.
For that a combined model that incorporates the full fertility
history of women would be more appropriate.
A model of this type would allow us to jointly model fertility
and sex selection and help address both questions.

The results here also lead to a set of broader important questions that
need to be addressed in future research.
First, to what extent are the improvements in health for girls
relative to boys the result of selection, longer spacing between
births, or changing son preference?
There is some early evidence that selection make it appear that
girls are healthier, even though the real cause is a combination
of sex selection and higher mortality together with recall error \citep{Portner2018a}. 
Understanding what role these factors play in better health for
girls is especially important when evaluating policies that 
aim to directly limit the use of sex selection rather than 
changing preferences or incentives.

Second, we need to understand how female labor force participation
interact with the use of sex selection.
Increased autonomy for women, through, for example, better opportunities
for working outside the home, has been suggested as a way to increase
women's status and thereby lowering the use of sex selection \citep{Das-Gupta2016}.
This may, however, be a double-edged sword.
On the one hand, it is an obvious benefit to the women in terms of 
bargaining power and it also increases the cost of repeated 
sex-selective abortions.
On the other hand, it may also further lower fertility, and everything 
else equal, that may lead to higher use of sex selection.

@@

This suggestion that an overlooked approach to addressing a biased
sex ratio may be to encourage higher female labor force participation.

>>>>

Parents' spacing between births has long served as a measure of son preference 
\citep{Leung1988}.
Before prenatal sex determination became available, the only recourse for 
parents who wanted a son---but did not yet have one---was to have the next 
birth sooner.
Son preference is therefore often associated with shorter spacing after the 
births of girls than boys 
\citep{Das1987,Rahman1993,Pong1994,Haughton1996,Arnold1997,Soest2012,Rossi2015}.
Shorter spacing is, in turn, associated with worse health outcomes for girls 
and mothers 
\citep{arnold98,Conde-Agudelo2000,Whitworth2002,Razzaque2005,Rutstein2005,Conde-Agudelo2006}.%
\footnote{
Parents are also more likely to cease childbearing after the birth of 
a son than after a daughter 
\citep{repetto72,ben-porath76b,Das1987,Arnold1997,arnold98,clark00,dreze01,filmer09,Basu2010,Altindag2016}.
}

@@

% [Why India?]
I apply the method to birth histories for Hindu women, using data from 
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2016. 
India is a particularly compelling case.
On the one hand, India has seen dramatic increases in the males-to-females ratio 
at birth over the last three decades as access to prenatal sex determination 
expanded 
\citep{das_gupta97,Sudha1999,Arnold2002,retherford03b,jha06}.%
\footnote{
India is not alone; both China and South Korea saw
significant changes in the sex ratio at birth over the same period 
\citep{Yi1993,park95}.
}
On the other hand, research suggests that son preference in 
India, when measured as ideally having more boys than girls, is decreasing 
over time and with higher education \citep{bhat03,pande07}.
% %
% \footnote{
% This measure of son preference is commonly used in the literature. 
% See, for example, \citet{clark00}, \citet{Jensen2009}, and \cite{Hu2015}.
% }

There have also been substantial changes in access and legality of
prenatal sex determination in India over the period covered.
Abortion has been legal in India since 1971 and still is, but the 
first reports of prenatal sex determination did
not appear until 1982--83 \citep{Sudha1999,bhat06,Grover2006}.
The number of clinics quickly increased, and knowledge about sex selection 
became widespread after a senior government official's wife aborted a 
fetus that turned out to be male \citep[p.\ 598]{Sudha1999}.
In 1994, the Central Government passed the Prenatal Diagnostic Techniques 
(PNDT) Act, making determining and communicating the sex of a fetus illegal.%
\footnote{
Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
The number of convictions has been low.
It took until January 2008 for the first state, Haryana, to reach five convictions.
Hence, private clinics apparently operate with little risk of legal action 
\citep{Sudha1999}.
Furthermore, there is little evidence that bans like this significantly
affect sex ratios \citep{Das-Gupta2016}.
}
% Hence, the data make it possible to show how the spacing between births and
% sex ratios have changed with the introduction of prenatal sex determination,
% and whether the ban affected the relationship between birth spacing and sex 
% ratios.

% [Results]
There are four main results.
First, there has been a general increase in the length of spacing between births
over the four decades covered by the data.
The exception is for first births, where the median duration has either remained
the same or slightly declined, although this hides a substantial compression of
the variation in spell length.
Second, consistent with the use of sex selection, the increase in duration is 
substantially longer for women who are most likely to use sex selection than
for those less likely to use sex selection.
Specifically, for the best educated women, those with no boys now has 
significantly longer spacing and a more male-biased sex ratio than those 
with boys.
Third, women with no education still follow the standard pattern of short 
spacing when they have girls and little evidence of the use of sex selection.
In other word, these women follow a strategy where they achieve a son through
higher fertility rather than use of sex selection.
Finally, sex ratios are more likely to decline within spells at lower parities, 
where the pressure to ensure a son is lower, and more likely to increase or 
remain consistently high for higher-order spells, where the pressure to ensure 
a son is high.

@@

As mentioned above, 10 to 20 percent of women gave birth less than nine months
after they were married, and the first spell, therefore, begins at the 
month of marriage.
The exception is for women married very young, where I use the month they 
turned 12.
The second and subsequent spells begin nine months after the previous birth 
because that is the earliest we should expect to observe a new birth.
A few women report births that occurred less than nine months 
after the previous birth; I drop those women from the sample as mentioned above.

All spells continue until either a child is born or censoring occurs.
Censoring can happen for three reasons:
the survey takes place;
sterilization of the woman or her husband;
or too few births are observed for the method to work.
For the first spell I set censoring at 72 months (six years) after marriage,
while for the second and third spells censoring is at 72 months and 
for the fourth spell 57 months after a birth can occur.
With nine month between the previous birth and the beginning of
the next spell, the maximum durations between births I consider are
6.75 years for the second and third spells and 5.5 years for the 
fourth spell.

I group spells into four periods based on the year of marriage
or year of the previous birth:
1972--1984, 1985--1994, 1995--2004, and 2005--2016.
The periods follow the main changes in availability and legality of 
prenatal sex determination discussed in the Introduction.
The allocation into periods is determined by when conception and therefore 
decisions on sex selection begin, even if no births are observed until 
nine months later.

The allocation into periods means that some spells cover two periods.
A couple may, for example, be married in 1984, but not have their first 
child until 1986.
That couple's first spell will be in the 1972--1984 period, even though 
most of the interval falls in the 1985--1994 period.
Hence, prenatal sex determination was likely available when some
children from spells that began in the 1972--1984 period were conceived,
which could result in evidence of sex-selective abortions even for this
period.
Similarly, a spell that started in the 1985--1994 period may have been partly 
or mostly under the PDNT act.
The effect is a downward bias in the differences between the periods.

@@

The area of residence is a dummy variable for the household living in
an urban area.
The cost of children is higher in urban areas than in rural areas, and 
there is easier access to prenatal sex determination in urban areas.
Both factors are likely to lead to higher use of sex-selective abortions
in urban than in rural areas.

@@

The predicted sex ratio is the weighted average over the women in the sample's 
individual predicted sex ratios, using the individual probabilities of a birth by 
the end of the spell as weights.
Each woman's predicted sex ratio is the weighted average of the predicted
percentage boys over the months in the spell calculated using equation 
(\ref{eq:probability_son}) and the probability of giving birth in each month 
as weights.%
\footnote{
Imagine a spell has two periods and that the estimated percentage boys for a
woman are 54\% and 66\% and that the likelihood of having a birth is 20\% and 40\%
(measured as the change in the survival curve; in other words, there is 
40\% chance that she does not have a birth by the end of this spell).
This woman's percentage boys is then $\frac{54*0.2+66*0.4}{0.2+0.4} = 62$.
}
The predicted sex ratio captures the percent boys that will be born to women 
in the sample once child bearing for that spell is over.

@@

The above results show that length of birth spacing increased 
more for women with only girls than for other sex compositions
as sex ratios became more male-biased, presumably because
of increased use of sex selection.
Neither the median spacing or the conditional survival curves,
however, can tell us whether sex selection vary \emph{within}
a spell or provide any information about the likelihood of
exiting a spell with a birth.
To understand both, I now turn to the standard survival curves
and predicted sex ratios by spell duration.

@@

The predicted sex ratio may change within a spell for, at
least, two reasons.
First, parents may stop using prenatal sex determination within
a spell once the duration from the previous birth becomes long enough.
The desire for short spacing could, for example, arise from 
economies of scale in childbearing and rearing that would be lost 
if children were born too far apart \citep[p 374]{schultz97}. 
In this case we should observe the predicted sex ratio decline
with spell length.

Second, if the degree of son preference vary across women within 
the sample but behavior is not affected by duration from the 
previous birth, we should observe an increasing share of births come from 
parents with stronger son preference as duration from the
previous birth increases.
Whereas the early births are a combination of births to
people with lower son preference and those with stronger son
preference who conceived a son in the first or second pregnancy,
later births are then predominately from those with stronger
son preference who have not yet conceived a son.
In this case we should see an increase in the sex ratio 
as duration increases.

@@

At the other extreme, women with eight or more years of education have
experienced an almost complete reversal of their spacing patterns.
Rather than having the next birth sooner as they mostly did before 
sex selection became available, women with either no or one son now 
have substantially longer spacing than if they have two or more sons.
These changes are likely due to increased use of sex selection 
rather than changes in son preference as shown by the increasingly
significant deviations in the sex ratio from the natural sex ratio.
This also suggests that banning prenatal sex determination has 
not been able to reverse the use, although it is possible that it
has slowed the rate of increase.

@@

The results here point to two important, interconnected, questions 
for future research:
What is the connection between falling fertility and the 
use of sex selection and is the use of sex selection increasing 
or decreasing over time?
It is clear that fertility has been declining over time in India,
although it is important to note that the decline might be
overstated exactly because of an increased use of sex selection and
the resulting longer spacing between births makes it less likely 
that we will observe births to women, even though they eventually
will happen.
The facts that biased sex ratio occur at earlier spells over
time and starting to show up for women with less education 
all indicate that lower fertility and use of sex selection are
related, but a more detailed analysis is needed.
Furthermore, although examining spells separately, as done here, is a useful
first step in understanding how sex selection changes over time,
it does not lend itself easily to examine the degree to which
sex selection changes over time.
For that a combined model that incorporates the full fertility
history of women would be more appropriate.
A model of this type would allow us to jointly model fertility
and sex selection and help address both questions.

The results here also lead to a set of broader questions that
need to be addressed in future research.
First, to what extent are the improvements in health for girls
relative to boys the result of selection, longer spacing between
births, or changing son preference?
There is some early evidence that selection make it appear that
girls are healthier, even though the real cause is a combination
of sex selection and higher mortality together with recall error \citep{Portner2018a}. 
Understanding what role these factors play in better health for
girls is especially important when evaluating policies that 
aim to directly limit the use of sex selection rather than 
changing preferences or incentives.

Second, we need to understand how female labor force participation
interact with the use of sex selection.
Increased autonomy for women, through, for example, better opportunities
for working outside the home, has been suggested as a way to increase
women's status and thereby lowering the use of sex selection \citep{Das-Gupta2016}.
This may, however, be a double-edged sword.
On the one hand, it is an obvious benefit to women who gain  
bargaining power and it also increases the cost of repeated 
sex-selective abortions.
On the other hand, it may further lower desired fertility, and 
everything else equal, that may lead to higher use of sex selection.
Understanding the trade-off between long-term benefits from improvements in
women's labor force participation and short-term costs from potential 
increases in sex selection is of paramount importance.

@@
